---
title: "daymet_narr_noaa_validation"
author: "Ben Barrett"
date: last-modified
date-format: long
format: html
editor: visual
---

**Comparing the Daymet and NOAA station, and NARR and NOAA station, daily measures for average temperature (TAVG) at the three NOAA stations in Chicago over 2017-2024.**

```{r}
#| echo: false
#| message: false
#| warning: false
# Loading the necessary packages
library(tidyverse)
library(ggpubr)
library(blandr)
library(tmap)
library(sf)
library(janitor)

# Importing the NOAA station, Daymet, and NARR temperature data
noaa_stations_daymet_narr <- read_csv("noaa_stations_daymet_narr.csv")
```

**Map of Chicago NOAA station observation sites.**

```{r}
#| echo: false
#| message: false
# Loading Chicago boundary shapefile
chicago <- st_read(dsn = "City_Boundary_20250407/geo_export_6217bba7-ddc8-4f38-b5cd-7ca9ba5edd5a.shp", quiet = TRUE)

# Loading NOAA stations
stations <- read_csv("noaa_stations.csv")
stations <- stations %>%
  select(name, lat, lon) %>%
  distinct()
stations$name[stations$name == "CHICAGO MIDWAY AIRPORT, IL US"] <- "Midway Airport"
stations$name[stations$name == "CHICAGO NORTHERLY ISLAND, IL US"] <- "Northerly Island"
stations$name[stations$name == "CHICAGO OHARE INTERNATIONAL AIRPORT, IL US"] <- "O'Hare Airport"

# Converting NOAA stations to shape object
stations <- st_as_sf(stations, coords = c('lon', 'lat'), crs = st_crs(chicago)$proj4string)

# Creating map
stations <- stations %>%
  mutate(x_offset = 0,
         y_offset = 0,
         x_offset = case_when(
           name == "Midway Airport" ~ 0.3,
           name == "Northerly Island" ~ 3.2,
           name == "O'Hare Airport" ~ -3.2,
           TRUE ~ x_offset),
         y_offset = case_when(
           name == "Midway Airport" ~ 0.75,
           name == "Northerly Island" ~ 0,
           name == "O'Hare Airport" ~ 0,
           TRUE ~ y_offset))
tm_shape(chicago) +
  tm_polygons() +
  tm_scale_bar(position = c(0.2, 0.1)) +
  tm_layout(frame = FALSE, inner.margins = c(0, 0.25, 0, 0.05)) +
  tm_shape(stations) +
  tm_dots(fill = "red", size = 0.3) +
  tm_text("name", size = 0.75, xmod = "x_offset", ymod = "y_offset")
```

**Number of observations/estimates (temperature measurements) by year and NOAA station.**

```{r}
#| echo: false
station_count <- noaa_stations_daymet_narr %>%
  group_by(name, year(date)) %>%
  count() %>%
  rename(year = `year(date)`)

station_count <- station_count %>%
  group_by(name) %>%
  group_modify(~ adorn_totals(.x))

print(station_count, n = 27)
```

**Violin plot of NOAA, Daymet, and NARR daily average temperature distributions.**

```{r}
#| echo: false
violin_data_noaa <- noaa_stations_daymet_narr %>%
  select(tavg_noaa) %>%
  rename(tavg = tavg_noaa) %>%
  mutate(Data = "NOAA")
violin_data_daymet <- noaa_stations_daymet_narr %>%
  select(tavg_daymet) %>%
  rename(tavg = tavg_daymet) %>%
  mutate(Data = "Daymet")
violin_data_narr <- noaa_stations_daymet_narr %>%
  select(tavg_narr) %>%
  rename(tavg = tavg_narr) %>%
  mutate(Data = "NARR")
violin_data <- rbind(violin_data_noaa, violin_data_daymet, violin_data_narr)
violin_data <- violin_data %>%
  mutate(Data = fct_relevel(Data, "Daymet", "NOAA", "NARR"))

ggplot(violin_data, aes(x = Data, y = tavg)) +
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.1, outliers = TRUE) +
  ylab("Daily Average Temperature (Celsius)") +
  xlab("") +
  labs(caption = "Chicago, Illinois; 1 January 2017 - 30 December 2024; N = 8514") +
  scale_y_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

**Two-sample Kolmogorov-Smirnov test for Daymet vs. NOAA daily average temperature.**

```{r}
#| echo: true
ks.test(noaa_stations_daymet_narr$tavg_daymet, noaa_stations_daymet_narr$tavg_noaa, alternative = "two.sided", exact = TRUE)
```

The p-value is not statistically significant (p = 0.10) and so the null hypothesis that Daymet and NOAA were drawn from the same distribution is not rejected.

**Two-sample Kolmogorov-Smirnov test for NARR vs. NOAA daily average temperature.**

```{r}
#| echo: true
ks.test(noaa_stations_daymet_narr$tavg_narr, noaa_stations_daymet_narr$tavg_noaa, alternative = "two.sided", exact = TRUE)
```

The p-value is statistically significant (p \< 0.01) and so the null hypothesis that NARR and NOAA were drawn from the same distribution is rejected.

**Scatterplot for Daymet vs. NOAA daily average temperature.**

```{r}
#| echo: false
ggplot(noaa_stations_daymet_narr, aes(x = tavg_noaa, y = tavg_daymet)) +
  geom_point(size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  xlab("NOAA Daily Average Temperature (Celsius)") +
  ylab("Daymet Daily Average Temperature (Celsius)") +
  labs(caption = "Chicago, Illinois; 1 January 2017 - 30 December 2024; N = 8514") +
  scale_x_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  stat_cor(method = "pearson", label.x = -28, r.digits = 3, p.accuracy = 0.01) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

**Bland-Altman plot for Daymet vs. NOAA daily average temperature.**

```{r}
#| echo: false
#| warning: false
p <- blandr.draw(noaa_stations_daymet_narr$tavg_noaa, noaa_stations_daymet_narr$tavg_daymet, ciDisplay = FALSE, ciShading = FALSE) +
  xlab("Mean of NOAA and Daymet Daily Average Temperature (Celsius)") +
  ylab("Difference between NOAA and Daymet\nDaily Average Temperature (Celsius)") +
  labs(caption = "Chicago, Illinois; 1 January 2017 - 30 December 2024; N = 8514", title = "") +
  scale_x_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-15, 15), breaks = seq(-15, 15, by = 5), expand = c(0, 0)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))

p_build <- ggplot_build(p)
p_build$data[[3]][["colour"]] <- "blue"
p_build$data[[4]][["colour"]] <- "red"
p_build$data[[5]][["colour"]] <- "red"
grid::grid.draw(ggplot_gtable(p_build))
```

**Scatterplot for NARR vs. NOAA daily average temperature.**

```{r}
#| echo: false
ggplot(noaa_stations_daymet_narr, aes(x = tavg_noaa, y = tavg_narr)) +
  geom_point(size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  xlab("NOAA Daily Average Temperature (Celsius)") +
  ylab("NARR Daily Average Temperature (Celsius)") +
  labs(caption = "Chicago, Illinois; 1 January 2017 - 30 December 2024; N = 8514") +
  scale_x_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  stat_cor(method = "pearson", label.x = -28, r.digits = 3, p.accuracy = 0.01) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

**Bland-Altman plot for NARR vs. NOAA daily average temperature.**

```{r}
#| echo: false
#| warning: false
p <- blandr.draw(noaa_stations_daymet_narr$tavg_noaa, noaa_stations_daymet_narr$tavg_narr, ciDisplay = FALSE, ciShading = FALSE) +
  xlab("Mean of NOAA and NARR Daily Average Temperature (Celsius)") +
  ylab("Difference between NOAA and NARR\nDaily Average Temperature (Celsius)") +
  labs(caption = "Chicago, Illinois; 1 January 2017 - 30 December 2024; N = 8514", title = "") +
  scale_x_continuous(limits = c(-30, 35), breaks = seq(-30, 35, by = 5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-15, 15), breaks = seq(-15, 15, by = 5), expand = c(0, 0)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"))

p_build <- ggplot_build(p)
p_build$data[[3]][["colour"]] <- "blue"
p_build$data[[4]][["colour"]] <- "red"
p_build$data[[5]][["colour"]] <- "red"
grid::grid.draw(ggplot_gtable(p_build))
```
